---
title: "Downloading NVE Grid Time Series data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading NVE Grid Time Series data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

The NVE Grid Time Series (GTS) API serves gridded environmental data for Norway
on a 1 km grid: precipitation, temperature, snow, wind, soil moisture,
groundwater and a pile of derived hydrological indices. You query it one grid
point at a time, over a date range, and get a JSON time series back.

`NveGTSRequesteR` handles the tedious parts: building the URLs, fetching them in
parallel, writing results to disk in batches so a long job cannot eat your RAM,
and picking up where it left off if the job dies halfway.

This vignette walks through a full job in pieces, so you can stop after the
part you actually need.

## 1. Setup

```{r}
library(NveGTSRequesteR)
library(future)
library(sf)
library(dplyr)
library(purrr)
```

Parallel fetching is opt-in: `download_nve_gts()` calls `furrr::future_map()`,
which runs sequentially unless you set a `future::plan()` first.

```{r}
plan(multisession, workers = parallelly::availableCores() - 1)  # availableCores() is the count to use: it respects cgroup/container/HPC limits that detectCores() ignores; subtract 1 to leave a core for your own session
```

These are network-bound requests, not CPU-bound work, so you can safely go
above your core count if the API keeps up — `workers = 12` on an 8-core machine
is not unreasonable. Do not go wild; you are hitting a public service.

## 2. Pick a layer

`VALID_GTS_LAYER_NAMES` is the list of short-codes the API accepts.
`gts_layers_full` carries the metadata for each one: full name, unit, time
resolution, no-data sentinel, and how far back the series goes.

```{r}
# Daily layers only (1440 minutes = 24 hours)
gts_layers_full |>
  keep(\(x) x$TimeResolutionInMinutes == 1440L) |>
  map_chr(\(x) x$Name)

# What does "tm" actually give me?
gts_layers_full |>
  detect(\(x) x$Name == "tm") |>
  str()
#> List of 8
#>  $ Name                   : chr "tm"
#>  $ Fullname               : chr "Døgntemperatur"
#>  $ FullnameEN             : chr "Daily mean temperature"
#>  $ NoDataValue            : int 65535
#>  $ RawUnit                : chr "Kelvin"
#>  $ HumanReadableUnit      : chr "Celcius"
#>  $ TimeResolutionInMinutes: int 1440
#>  $ FirstDateInTimeSerie   : chr "1957-01-01"
```

`FullnameEN` is this package's own addition — the API returns only the
Norwegian `Fullname`. Layer names are terse, so it spells out the jargon:

```{r}
gts_layers_full |>
  keep(\(x) x$Name %in% c("swe", "qtt", "sca")) |>
  map_chr(\(x) x$FullnameEN)
#> [1] "Snow water equivalent (depth of water the snowpack would give if fully melted)"
#> [2] "Snow-covered area (share of the grid cell lying under snow)"
#> [3] "Water input to the ground (rain plus snowmelt reaching the surface)"
```

Check `FirstDateInTimeSerie` before you set your date range. Most of the
seNorge-derived layers start in 1957; the model-derived ones often start in
2010 or 2018, and asking for earlier dates gets you no-data values rather than
an error.

## 3. Get your coordinates

`download_nve_gts()` wants a data frame with numeric `X` and `Y` columns in
**UTM33N metres**. That is the only thing the API speaks. There are two normal
ways to arrive there.

### 3a. A handful of known points

```{r}
coords <- data.frame(
  X = c(269390, 270000),
  Y = c(6653390, 6654000)
)
```

### 3b. A grid covering a polygon

This is the common case: you have an area — a catchment, a municipality, a
precipitation region — and you want every grid cell inside it.

```{r}
area <- read_sf("~/my_area.geojson")

st_crs(area)   # what is the file actually in? Never assume
st_bbox(area)  # sanity check: UTM33N northings over Norway run roughly 6.45e6–7.95e6
```

Most Norwegian datasets are already UTM33N, so this transform is often a no-op —
but it is worth doing anyway. Data from the far north is frequently distributed
in UTM35N, and anything sourced internationally tends to arrive in WGS84
lat/long or in whatever UTM zone its own country uses.

```{r}
# `crs =` is the TARGET; the source CRS is read off the object itself.
# EPSG:25833 = ETRS89 / UTM zone 33N. ETRS89 UTM codes are 25800 + zone number,
# so zone 32 = 25832, zone 33 = 25833, zone 35 = 25835.
# (For comparison: WGS84 UTM north is 32600 + zone, and plain WGS84 lat/long
#  is 4326. NVE only accepts 25833 coordinates, whatever you started from.)
area_utm33 <- st_transform(area, crs = 25833)
```

Now lay a regular grid over it. `cellsize = 1000` matches the native 1 km GTS
grid — going finer just queries the same cells repeatedly.

```{r}
grid_points <- st_make_grid(area_utm33, cellsize = 1000, what = "centers") |>
  st_intersection(area_utm33) |>   # keep only centres inside the polygon
  st_coordinates() |>
  as.data.frame() |>
  setNames(c("X", "Y"))

nrow(grid_points)  # this is your request count — look at it before downloading
```

That last line matters. One point × one layer × one date range is one HTTP
request. A mid-sized municipality is a few thousand; half of Norway is a few
hundred thousand.

## 4. Download

**Before you call it, look at `batch_size`.** The default of `100` points per
batch is tuned for the large case — pulling an entire region, tens of thousands
of points, where the batch is a memory ceiling and a resume checkpoint. That is
not the usual job. Downloading twelve points for a few sites is far more common,
and there the default leaves performance on the table: work is parallelised
*within* a batch, so a batch smaller than your worker pool idles workers, and
every batch boundary costs a disk write.

Rule of thumb:

- **Small job** (up to a few hundred points): set `batch_size` at or above your
  point count so the whole thing runs as one parallel sweep.
- **Large job** (thousands+): keep it near the default, and make it a multiple
  of your worker count. Raise it if you have RAM to spare, lower it if each
  response is huge (long date ranges, hourly layers).

```{r}
folder <- download_nve_gts(
  name            = "my_area",
  coordinate_list = grid_points,
  first_date      = "2023-01-01",
  last_date       = "2023-12-31",
  layer           = "tm",
  batch_size      = 100
)

folder
#> "my_area_tm_2023-01-01_2023-12-31_x269-412_y6653-6698_4812pts"
```

The folder is created relative to your working directory and named after the
request itself: layer, date range, the bounding box of your points rounded to
whole kilometres, and the point count. Inside it you get one `.rds` per batch,
plus a manifest.

Three properties worth knowing:

- **Resumable.** Batches whose `.rds` already exists are skipped. If the run
  dies — and a region-sized run will eventually meet a flaky network — rerun
  the identical call and it picks up where it stopped.
- **It will not resume the wrong thing.** The manifest inside the folder stores
  the full coordinate vector along with the dates, layer and batch size. A call
  that differs in any of those is refused with an error naming the field, so
  changing your date range and rerunning cannot silently give you a folder
  half-full of last week's results. To genuinely start over in the same place,
  clear it with `gts_folder_reset()`.
- **Failures are silent.** `request_nve_gts()` returns `NULL` on any HTTP error
  or network failure rather than aborting the whole run, so one bad point does
  not kill a six-hour job. You check for these after loading.

That second point is why the folder name is only a label and not the safety
mechanism. A bounding box describes an *extent*, not a *set of points*: twelve
scattered sites and a dense grid clipped from the same polygon produce the same
name. The manifest is what tells them apart.

## 5. Load the results back

Pass the folder name straight through — no need to reconstruct it:

```{r}
results <- load_and_bind_gts_from_folder(folder)
```

If you are coming back to a folder in a later session, rebuild the name with
`name_folder()` using the same arguments you downloaded with:

```{r}
folder <- name_folder("my_area", "tm", grid_points, "2023-01-01", "2023-12-31")
```

This returns a **flat list**, one element per grid point, in the same order as
the coordinates you passed in — but not a data frame. Each element is the
parsed JSON response:

```{r}
str(results[[1]])
#> List of 11
#>  $ Theme            : chr "tm"
#>  $ FullName         : chr "Døgntemperatur"
#>  $ NoDataValue      : int 65535
#>  $ X                : int 269390
#>  $ Y                : int 6653390
#>  $ StartDate        : chr "01.01.2023 06:00:00"
#>  $ EndDate          : chr "31.12.2023 06:00:00"
#>  $ PrognoseStartDate: NULL
#>  $ Unit             : chr "Celcius"
#>  $ TimeResolution   : int 1440
#>  $ Altitude         : int 163
#>  $ Data             : List of 365
```

Note `Unit`: the API has already applied the scaling from `RawUnit` to
`HumanReadableUnit` for you. `tm` comes back in Celsius, not Kelvin; `sd` in
centimetres, not millimetres. `Data` is a bare array with no timestamps — the
time axis is implied by `StartDate` plus `TimeResolution` minutes per step.

The request that produced the data travels with it, so you can always tell what
a loaded object actually is:

```{r}
attr(results, "gts_manifest")$first_date
attr(results, "gts_manifest")$layer
```

Check for failed requests before you go further. Don't hand-count nulls — ask
the question directly:

```{r}
every(results, \(x) !is.null(x))  # TRUE if all requests came back

results <- compact(results)  # drop the failures, if you accept losing them
```

If some failed, the cheaper fix is usually to delete the affected batch files
and rerun the same `download_nve_gts()` call — the resume logic refetches only
what is missing.

## 6. Rectangle it into a tibble

The package stops at the list. Turning that into a tidy frame is a few lines,
and worth writing once:

```{r}
gts_to_tibble <- function(x) {
  start <- as.POSIXct(x$StartDate, format = "%d.%m.%Y %H:%M:%S", tz = "Europe/Oslo")
  values <- unlist(x$Data)

  tibble::tibble(
    X        = x$X,
    Y        = x$Y,
    altitude = x$Altitude,
    layer    = x$Theme,
    unit     = x$Unit,
    time     = start + (seq_along(values) - 1) * x$TimeResolution * 60,
    value    = ifelse(values == x$NoDataValue, NA_real_, values)
  )
}

gts <- results |>
  map(gts_to_tibble) |>
  list_rbind()

gts
#> # A tibble: 4,380 × 7
#>         X       Y altitude layer unit    time                value
#>     <int>   <int>    <int> <chr> <chr>   <dttm>              <dbl>
#>  1 269390 6653390      163 tm    Celcius 2023-01-01 06:00:00   0.1
#>  2 269390 6653390      163 tm    Celcius 2023-01-02 06:00:00  -5
#>  ...
```

The `NoDataValue` check is the part people forget. `65535` is a sentinel, not a
temperature — leave it in and it will quietly wreck every mean you compute.

From here it is ordinary dplyr:

```{r}
gts |>
  group_by(X, Y) |>
  summarise(mean_temp = mean(value, na.rm = TRUE), .groups = "drop")
```

And back to spatial when you need it — the coordinates are still UTM33N, so
that is what you tell `st_as_sf()`:

```{r}
gts_sf <- gts |>
  group_by(X, Y) |>
  summarise(mean_temp = mean(value, na.rm = TRUE), .groups = "drop") |>
  st_as_sf(coords = c("X", "Y"), crs = 25833)  # `crs =` here DECLARES the CRS the numbers are already in — it does not convert
```

## 7. Cleaning up

`.rds` batches for a large region add up. When you are done:

```{r}
gts_folder_reset(folder)
```

It asks for confirmation before deleting. This also removes the manifest, which
is what frees the folder up to be reused for a different date range or a
different set of points.
